refnx -- Neutron and X-ray reflectometry analysis in Python

### 2018-07-25 paper/manuscript.tex
PAPER:method

The \Objective\ class uses the \ReflectModel\ and a dataset, \DataD, to calculate $\chi^2$, log-likelihood (Equation~\ref{eqn:2}), log-prior, residuals, and the generative model.
The \DataD\ object has \emph{x, y, y\_err, x\_err} attributes to represent $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$. The \DataD\ object can be constructed from a tuple of arrays representing the data, or by reading a three or four column plain-text datafile. A three column dataset represents $Q$ (\si{\per\angstrom}), $R$, $\mathrm{d}R$ (1 standard deviation). A four column dataset represents $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$ (\si{\per\angstrom}). Subclassing \DataD\ would allow other formats to be read. One example of this could be an wavelength dispersive file using $\Omega$, $\lambda$ data instead of $Q$; such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers. In such a case \ReflectModel\ could be subclassed to make full use of this energy dispersive information.
An \Objective\ can be given a \Transform\ object to permit fitting as $\log_{10} R$ vs $Q$, $RQ^4$ vs $Q$; the default (no \Transform) is $R$ vs $Q$. Several \Objective\ can be combined to form a \GlobalObjective\ for co-refinement. The object-oriented nature allows \Parameter, \Component\ to be reused in in different \Objective, thereby facilitating co-refinement. For a comprehensive demonstration of multiple contrast co-refinement, see the annotated notebook in the supporting information.

The \Objective\ statistics are used directly by the \CurveFitter\ class to perform least-square fitting with the functionality provided by the \SciPy\ package (DifferentialEvolution, Levenberg--Marquardt, LBFGSB - Limited Broyden--Fletcher--Goldfarb--Shanno with bounds).
\CurveFitter\ can also perform Bayesian Markov Chain Monte Carlo (MCMC) sampling of the system, examining the posterior probability distribution of the parameters, Equation~\ref{eqn:1}. The posterior distribution is proportional to the product of the prior probability and the likelihood (or the sum of the log-probabilities):
%
\begin{gather} 
\label{eqn:1}\ p(\theta | D, I) \propto p(\theta | I)\times p(D | \theta, I)\\
p(D | \theta, I) = -\frac{1}{2} \sum \left[\left(\frac{y_n - model} {\sigma_n}\right)^2 + \log(\sigma_n^2)\right]\label{eqn:2}
\end{gather}
%
The prior distribution, $p(\theta | I)$, is the probability distribution function for a parameter, $\theta$, given preexisting knowledge of the system, $I$, as outlined above.
The likelihood (Equation~\ref{eqn:2}), $p(D | \theta, I)$, is the probability of the observed data, $D$, given the model parameters and other prior information. It is calculated from the measured data, $y_n$ (with uncertainties $\sigma_n$), and the generative model.

The posterior distribution, $p(\theta | I, D)$, is the probability distribution function of the parameters, given the data and prior knowledge about the system.  The posterior probability is examined by encoding the likelihood and prior distributions and using the \emcee\ and \ptemcee\ packages to perform affine invariant ensemble sampling \cite{emcee, ptemcee}. \ptemcee\ is a fork of \emcee\ that implements parallel tempering for characterisation of multi-modal probability distributions; different modes can be traversed by chain populations at higher `temperatures', while individual modes are efficiently explored by chains at lower `temperatures'.

The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective\ overriding the log-likelihood method.

@@

To make it easy to create quick models, and to ease the learning curve, a browser-based graphical user interface (GUI) \cite{ipywidgets} can be instantiated for fitting within a \Jupyter\ notebook, Figure~\ref{fig:gui}. This GUI uses slab based models for analysing a single dataset, and can be used as a stepping point for building more advanced models via the `To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}.

>>>>

The \Objective\ class uses the \ReflectModel\ and a dataset, \DataD, to calculate $\chi^2$, log-likelihood (Equation~\ref{eqn:2}), log-prior, residuals, and the generative model.
The \DataD\ object has \emph{x, y, y\_err, x\_err} attributes to represent $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$. The \DataD\ object reads a three or four column plain-text datafile. A three column dataset represents $Q$ (\si{\per\angstrom}), $R$, $\mathrm{d}R$ (1 standard deviation). A four column dataset represents $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$ (\si{\per\angstrom}). Subclassing \DataD\ would allow other formats to be read. One example of this could be an wavelength dispersive file using $\Omega$, $\lambda$ data instead of $Q$; such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers. In such a case \ReflectModel\ could be subclassed to make full use of this energy dispersive information.
An \Objective\ can be given a \Transform\ object to permit fitting as $\log_{10} R$ vs $Q$, $RQ^4$ vs $Q$; the default (no \Transform) is $R$ vs $Q$. Several \Objective\ can be combined to form a \GlobalObjective\ for co-refinement. The object-oriented nature allows \Parameter, \Component\ to be reused in in different \Objective, thereby facilitating co-refinement. For a comprehensive demonstration of multiple contrast co-refinement, see the annotated notebook in the supporting information.

The \Objective\ statistics are used directly by the \CurveFitter\ class to perform least-square fitting with the functionality provided by the \SciPy\ package (DifferentialEvolution, Levenberg--Marquardt, LBFGSB - Limited Broyden--Fletcher--Goldfarb--Shanno with bounds).
\CurveFitter\ can also perform Bayesian Markov Chain Monte Carlo (MCMC) sampling of the system, examining the posterior probability distribution of the parameters, Equation~\ref{eqn:1}. The posterior distribution is proportional to the product of the prior probability and the likelihood (or the sum of the log-probabilities):
%
\begin{gather} 
\label{eqn:1}\ p(\theta | D, I) \propto p(\theta | I)\times p(D | \theta, I)\\
p(D | \theta, I) = -\frac{1}{2} \sum \left[\left(\frac{y_n - model} {\sigma_n}\right)^2 + \log(\sigma_n^2)\right]\label{eqn:2}
\end{gather}
%
The prior, $p(\theta | I)$, is the probability distribution function for a parameter, $\theta$, given preexisting knowledge of the system, $I$, as outlined above.
The likelihood (Equation~\ref{eqn:2}), $p(D | \theta, I)$, is the probability of the observed data, $D$, given the model parameters and other prior information. It is calculated from the measured data, $y_n$ (with uncertainties $\sigma_n$), and the generative model. The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective\ overriding the log-likelihood method.

The posterior probability, $p(\theta | I, D)$, is examined by encoding the likelihood and prior distributions and using the \emcee\ and \ptemcee\ packages to perform affine invariant ensemble sampling \cite{emcee, ptemcee}. \ptemcee\ is a fork of \emcee\ that implements parallel tempering for characterisation of multi-modal probability distributions; different modes can be traversed by chain populations at higher `temperatures', while individual modes are efficiently explored by chains at lower `temperatures'.

@@

To make it easy to create quick models, and to ease the learning curve, a browser-based graphical user interface (GUI) \cite{ipywidgets} can be used to fit within a \Jupyter\ notebook, Figure~\ref{fig:gui}. This GUI uses slab based models for analysing a single dataset, and can be used as a stepping point for building more advanced models via the `To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}.
